Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-02 paper.tex
edits from yan

% Recombination is hard to detect
Detecting recombination in SARS-CoV-2 is difficult, and identifying new
recombinant strains is a time-consuming, manual
process~\citep{Smith2023-identifying}.
% (e.g., \url{https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/}).
% This resource shows the typical workflow for manually identifying recombinants that the pango team uses: https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/
% Regarding bias, not peer-reviewed but in the official pango designation post for XAC, you can see that me (Katherine) and the authors had a multi-month long discussion on how many breakpoints there should be and where. Reproducibility was difficult: https://github.com/cov-lineages/pango-designation/issues/590
Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of characteristic mutations from different lineages
\citep[e.g.,][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}
and as a result can only identify inter-lineage recombination events.
\cite{Turakhia2022-it} presented a
phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
over ten million SARS-CoV-2 genomes, without pre-specifying sets of characteristic
mutations.
RIPPLES finds candidate recombinants by using an existing phylogeny
(built assuming no recombination) and finding potential recombinants
by  scanning for branches containing many mutations.
It then determines if these candidates would be better explained by
recombination by exhaustively breaking each
sequence into segments and attempting to find more parsimonious placements for
each segment on the phylogeny. If such placements are found, the sequence is
identified as a putative recombinant. Although it enables rapid searches for
genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
accounts for only mutations, treats recombinants \textit{post hoc}, and is an
incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
As noted by the authors, a \textit{post hoc} treatment of recombination is
possible when recombinant lineages are rare and leave few descendants. However,
the proliferation of recombinant lineages is making this increasingly
untenable; for example, more than half of the sequences sampled in February 2023
are from the recombinant strain XBB and its descendants~\citep{Chen2022-pz}. This also means that
future evolution of SARS-CoV-2 is likely to involve multiple sequential
recombination events on top of existing recombinant lineages, creating a highly
reticulated genealogy.

@@

We encode ambiguous nucleotide letters (i.e.,
not A, C, G, T, or a gap) in the pairwise genome alignments as missing data
(N). Problematic bases in the alignments, which had two or more Ns or
gaps within a distance of seven bases, are masked as missing data following
the approach used in the ``faToVCF'' tool used by
UShER~\citep{Turakhia2021-ur}.
Sites that are masked by this process are
treated as missing data by the LS HMM (Section~\ref{sec:ls}).
In addition, we exclude 481 problematic sites flagged as prone to
sequencing errors or as highly homoplasic entirely
(\url{https://github.com/W-L/ProblematicSites_SARS-CoV2/},
accessed 2022-09-22).

@@

\bibliographystyle{abbrvnat}
\bibliography{paper}

>>>>

% Recombination is hard to detect
Detecting recombination in SARS-CoV-2 is difficult, and identifying new
recombinant strains is a time-consuming, manual process, subject
to biases [FIXME: have we a citation for how tedious this is? Also what
are the biases?]
% This resource shows the typical workflow for manually identifying recombinants that the pango team uses: https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/
% Regarding bias, not peer-reviewed but in the official pango designation post for XAC, you can see that me (Katherine) and the authors had a multi-month long discussion on how many breakpoints there should be and where. Reproducibility was difficult: https://github.com/cov-lineages/pango-designation/issues/590
Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of characteristic mutations from different lineages
\citep[e.g.,][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}
and as a result can only identify inter-lineage recombination events.
\cite{Turakhia2022-it} presented a
phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
over ten million SARS-CoV-2 genomes, without pre-specifying sets of characteristic
mutations.
RIPPLES finds candidate recombinants by using an existing phylogeny
(built assuming no recombination) and finding potential recombinants
by  scanning for branches containing many mutations.
It then determines if these candidates would be better explained by
recombination by exhaustively breaking each
sequence into segments and attempting to find more parsimonious placements for
each segment on the phylogeny. If such placements are found, the sequence is
identified as a putative recombinant. Although it enables rapid searches for
genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
accounts for only mutations, treats recombinants \textit{post hoc}, and is an
incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
As noted by the authors, a \textit{post hoc} treatment of recombination is
possible when recombinant lineages are rare and leave few descendants. However,
the proliferation of recombinant lineages is making this increasingly
untenable; for example, more than half of the sequences sampled in February 2023
are from the recombinant strain XBB and its descendants~\citep{Chen2022-pz}. This also means that
future evolution of SARS-CoV-2 is likely to involve multiple sequential
recombination events on top of existing recombinant lineages, creating a highly
reticulated genealogy.

@@

We encode ambiguous nucleotide letters (i.e.,
not A, C, G, T, or a gap) in the pairwise genome alignments as missing data
(N). Problematic bases in the alignments, which had two or more Ns or
gaps within a distance of seven bases, are masked as missing data following
the approach used in the ``faToVCF'' tool used by
UShER~\citep{Turakhia2021-ur}.
Sites that are masked by this process are
treated as missing data by the LS HMM (Section~\ref{sec:ls}).
In addition, we exclude 481 problematic sites flagged as prone to
sequencing errors or as highly homoplasic
(\url{https://github.com/W-L/ProblematicSites_SARS-CoV2/},
accessed 2022-09-22) entirely.

@@

\bibliographystyle{plainnat}
\bibliography{paper}
